---
title: It is currently possible to permanently delete files in R with both file.copy()
  and file.create()
author: Pasi Haapakorva
date: '2019-04-25'
categories:
  - R
tags:
  - r
slug: it-is-currently-possible-to-permanently-delete-files-in-r-with-both-file-copy-and-file-create
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, eval = FALSE, warning = FALSE)
```
After an incident at work that involved deleting an entire project I discovered a dangerous bug (or feature)  in base R functions `file.copy()` and `file.create()`.

Let's first create a directory and a file inside it.

```{r}
dir.create("foo")
cat(file = "foo/bar", "baz\n")
readLines("foo/bar")
#> [1] "baz"
```
We'll then permanently delete the file contents.
```{r}
file.copy("foo", ".", recursive = TRUE)
#> [1] TRUE
readLines("foo/bar")
#> character(0)
```
```{r echo=FALSE}
unlink("foo")
```
What happens?

Looking at `file.copy()` code, we can see that `from` and `to` are checked in an `if` statement `if (recursive && to %in% from)`, but because `"foo"` is not `"."`, the statement is not triggered. If it had been, the function would have stopped `stop("attempt to copy a directory to itself")`.

It looks like (there's a call to C code) `file.copy()` then goes on to read the files in `from` and `file.create()` them in `to` directory. `file.create()` has another interesting property: it doesn't check if the file already exists. It goes on to create the file, which already exists, and removes all its contents.

## A fix

Obviously, `from` needs to be checked if it's the same as `to` and in the current if statement this is attempted. However it only checks if these two strings are exactly the same, but copying a directory to its own parent directory doesn't trigger the condition. We need to check if `to` is in the parent directory of `from`.

```{r}
if (recursive && (to %in% from || 
      normalizePath(dirname(normalizePath(from))) == normalizePath(to)))
```
Here `normalizePath()` is used because it, according to its help file, "Convert[s] file paths to canonical form for the platform, to display them in a user-understandable form and so that _relative and absolute paths can be compared._" `dirname()` outputs different path form so both input and output of `dirname()` need to be put through `normalizePath()`.

Another option would be to see if `from %in% list.files(to)` but I think this could be slow if `to` is a large directory.

If we use the if statement with `normalizePath()` in an otherwise identical function called `file.copy2()`, we can't copy the directory to itself.

```{r}
file.copy2("foo", ".", recursive = TRUE)
#> Error in file.copy2("foo", ".", recursive = TRUE): attempt to copy a directory to itself
```






